A sheaf--function dictionary for algebraic tori over local fields

### 2013-10-04 shanghai.tex
Working on Section 7

\newcommand{\FF}{{\mathbb{F}}}
\newcommand{\ZZ}{{\mathbb{Z}}}
\newcommand{\NN}{{\mathbb{N}}}
\newcommand{\CC}{{\mathbb{C}}}
\newcommand{\QQ}{{\mathbb{Q}}}
\newcommand{\RR}{{\mathbb{R}}}
\newcommand{\EE}{\mathbb{\bar Q}_\ell}
\newcommand{\OK}{\mathcal{O}_K}
\newcommand{\pK}{\mathfrak{p}_K}
\newcommand{\OKp}{\mathcal{O}_{K'}}
\newcommand{\pKp}{\mathfrak{p}_{K'}}
\newcommand{\OL}{\mathcal{O}_L}
\newcommand{\OO}[1]{\mathcal{O}_{#1}}
\newcommand{\Zp}{\mathbb{Z}_p}
\newcommand{\Qp}{\mathbb{Q}_p}
%\newcommand{\Fp}{\mathbb{F}_p}
\newcommand{\bFq}{\bar{k}}
\newcommand{\Fq}{k}
\newcommand{\Fqm}{k_m}
%\newcommand{\bFp}{{\mathbb{\bar F}_p}}
\newcommand{\WW}{\mathbb{W}}

\newcommand{\EEx}{\EE^\times}

@@

\begin{proposition}\label{prop:connected}
  If $G$ is a connected commutative algebraic group over $\Fq$ then 
  the full subcategories $\QCf(G) \hookrightarrow \QCb(G) \hookrightarrow \QC(G)$ 
  appearing in Sections~\ref{sec:bounded} and \ref{sec:finite} are equivalences 
  and
  \[
  \trFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
  is an isomorphism of groups.
\end{proposition}

@@

  Let $g$ be any $\Fq$-rational point on $G$ 
  and let $\bg$ be the geometric point on $G$ lying above $g$ 
  Since $G$ is connected, the geometric point $\bg$ determines
  an equivalence between the category of $\ell$-adic Weil local systems on $G$ and
  $\ell$-adic representations of $\W(G,\bg)$ \cite{deligne:80a}*{1.1.12}.  \David{We should include a definition for $\W(G,\bg)$, perhaps in an intro paragraph for the section}
  Now let $(\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$ 
  and let $\rho : \W(G, \bg) \to \EEx$ be the character determined by $(\gqcs{L},\phi)$. 
  The $\Fq$-rational point $g$ under the geometric point $\bg$ determines a splitting
  $\Weil{}\to \W(G,\bg)$ of $\W(G,\bg)\to \Weil{}$. 
  In this way the Weil local system $(\gqcs{L},\phi)$ determines an $\ell$-adic character $\rho_g : \Weil{} \to \EEx$.
  %\[
  %\begin{tikzcd}
  %1 \rar & \pi_1(\bG, \bg) \rar &  \arrow{d}{\rho} \W(G,\bg) \rar
  %& \ar[swap,bend right, dotted]{l}{g} \arrow[dotted]{dl}{\rho_g} \Weil{} \rar & 1\\
  %&&\EEx&&
  %\end{tikzcd}
  %\]
  This character is given by the trace of Frobenius of $\qcs{L}$ at $g$, as defined in Section~\ref{sec:Frob}:
  \[
  \rho_g(\Frob{}) =  \trFrob{\qcs{L}}(g).
  \]
  \David{It's not clear to me whether we've just proven this claim or if we're citing it from somewhere}

  On the other hand, we have already seen that $\trFrob{\qcs{L}} : G(\Fq) \to \EEx$
  is a group homomorphism. Since $G$ is an algebraic group over $\Fq$, it is a
  variety over $\Fq$ and thus $G(\Fq)$ is finite.
  Therefore $\trFrob{\qcs{L}}(g) = \rho_g(\Frob{})$ is a root of unity
  for every $g\in G(\Fq)$.  Since $\Weil{}$ is generated by
  $\Frob{}$ and $\rho_g : \Weil{} \to \EEx$ is
  a character, it follows that the image of $\rho_g$ is a finite group.
  Thus, $\rho_g$ extends to an $\ell$-adic character of $\Gal(\bFq/\Fq)$, 
  denoted by the same symbol, below.
  \[
  \begin{tikzcd}
  1 \rar & \ar[equal]{d} \pi_1(\bG, \bg) \rar & \W(G,\bg) \rar \dar & \Weil{} \arrow[bend left]{rr}{\rho_g} \rar \dar & 1 & \EEx\\
  1 \rar &  \pi_1(\bG, \bg) \rar & \pi_1(G,\bg) \rar & \Gal(\bFq/\Fq) \arrow[dotted]{rru} \rar & 1 &
  \end{tikzcd}
  \]

  We may now lift the $\ell$-adic character $\rho_g : \Gal(\bFq/\Fq) \to \EEx$
  to an $\ell$-adic character $\pi_1(G,\bg) \to \EEx$ using the canonical topological group homomorphism
  $\pi_1(G,\bg) \to \Gal(\bFq/\Fq)$ determined by the structure map for $G$. But $\bg$ also
  determines an equivalence between the category of $\ell$-adic
  representations of $\pi_1(G,\bg)$ and $\ell$-adic local systems on $G$. Let
  $\qcs{L}_0$ be a local system on $G$ in the isomorphism class
  determined by this $\ell$-adic character of $\pi_1(G,\bg)$.
  Then $b_G^*\qcs{L}_0 \iso \gqcs{L}$.
  Since $b_G^*$ is full and faithful (again, see \cite{deligne-katz:SGA7.2}*{Expos\'e XIII}
  or even \cite{beilinson-bernstein-deligne:81a}*{Prop. 5.2.1}),
  $
  b_{G\times G}^* : \Hom(m^*\qcs{L}_0,\qcs{L}_0\boxtimes\qcs{L}_0) \to \Hom({\bar m}^*\gqcs{L},\gqcs{L}\boxtimes\gqcs{L})
  $ 
  is a bijection 
  (hom taken in the categories on constructible $\ell$-adic sheaves on 
  $G\times G$ and ${\bar G}\times {\bar G}$ respectively, 
  in which $\ell$-adic local systems sit as full subcategories). 
  Let $\mu_0 : m^*\qcs{L}_0 \to \qcs{L}_0\boxtimes\qcs{L}_0$ be the isomorphism matching 
  $\mu : {\bar m}^*\gqcs{L} \to \gqcs{L}\boxtimes\gqcs{L}$, 
  the latter appearing in the definition of $\qcs{L}$. 
  Then, as in Section~\ref{sec:bounded}, $(\qcs{L}_0,\mu_0)$ is an object in $\QCb(G)$ 
  and $\qcs{L} \ceq (\gqcs{L},\mu,\phi)$ is isomorphic to $(b_G^*\qcs{L}_0,b_{G\times G}^*\mu_0)$ in $\QC(G)$.
  Thus, the full and faithful functor $B_G^* : \QCb(G) \to \QC(G)$ from Section~\ref{sec:bounded}
  is also essentially surjective, hence an equivalence.
  %(when $G$ is a connected, commutative algebraic group over $\Fq$).

@@

  To show that $L_G(f,\psi) \iso (\qcs{L}_0,\mu_0)$ we simply apply the sheaf--function dictionary
  as follows \citelist{\cite{laumon:87a}*{1.1.3} \cite{deligne:SGA4.5}*{Sommes trig.}}.
  First, recall that $L_G(f,\psi) \ceq (L(f,\psi), \mu(f,\psi))$.
  Let $k'/k$ be any degree-$n$ extension and let $\Nm : k'\to k$ be the norm map. 
  By \cite{laumon:87a}*{1.1.3.3}, 
  \[
  \trFrob{(G_{k'}\hookrightarrow G)^*L(f,\psi)} = \trFrob{\qcs{L}} \circ \Nm;
  \]
  on the other hand, 
  \[
  \trFrob{\qcs{L}} \circ \Nm = \trFrob{(G_{k'}\hookrightarrow G)^*\qcs{L}},
  \]
   by Proposition~\ref{prop:basechange}.
  %Using notation from \cite{laumon:87a}*{1.1}, this means $t_{\bullet L(f,\psi)} = t_{\bullet \qcs{L}_0}$.
  Using \cite{laumon:87a}*{Th\'eor\`eme~1.1.2}, it follows that  $L(f,\psi) \iso \qcs{L}_0$ in $\Loc(G)$.
  It is now clear that $(L(f,\psi),\mu(f,\psi)) \iso (\qcs{L}_0,\mu_0)$ in $\QCb(G)$.
  This completes the proof that $L_G : \QCf(G) \to \QC(G)$ is essentially surjective.
  Since we have already seen (in Section~\ref{sec:finite}) that $L_G$ is full and faithful, 
  it follows that $L_G$ is an equivalence.

  To finish the proof of Proposition~\ref{prop:connected}, it suffices to observe that the group homomorphism
  $
  \trFrob{G} : \QCiso{G} \to G(\Fq)^*
  $
  is surjective because each character $\chi \in G(\Fq)^*$ determines a finite quasicharacter
  sheaf $(\Lang,\chi^{-1})$ such that the trace of Frobenius of the quasicharacter sheaf
  $B_G^* (L_G(\Lang,\chi^{-1}))$ is $\chi$.
\end{proof}

>>>>

\newcommand{\FF}{{\mathbb{F}}}
\newcommand{\ZZ}{{\mathbb{Z}}}
\newcommand{\NN}{{\mathbb{N}}}
\newcommand{\CC}{{\mathbb{C}}}
\newcommand{\QQ}{{\mathbb{Q}}}
\newcommand{\RR}{{\mathbb{R}}}
\newcommand{\EE}{\mathbb{\bar Q}_\ell}
\newcommand{\Zl}{\mathbb{\bar Z}_\ell}
\newcommand{\OK}{\mathcal{O}_K}
\newcommand{\pK}{\mathfrak{p}_K}
\newcommand{\OKp}{\mathcal{O}_{K'}}
\newcommand{\pKp}{\mathfrak{p}_{K'}}
\newcommand{\OL}{\mathcal{O}_L}
\newcommand{\OO}[1]{\mathcal{O}_{#1}}
\newcommand{\Zp}{\mathbb{Z}_p}
\newcommand{\Qp}{\mathbb{Q}_p}
%\newcommand{\Fp}{\mathbb{F}_p}
\newcommand{\bFq}{\bar{k}}
\newcommand{\Fq}{k}
\newcommand{\Fqm}{k_m}
%\newcommand{\bFp}{{\mathbb{\bar F}_p}}
\newcommand{\WW}{\mathbb{W}}

\newcommand{\EEx}{\EE^\times}
\newcommand{\Zlx}{\mathbb{\bar Z}_\ell^\times}

@@

When $G$ is connected, we may apply the classic sheaf--function dictionary.  In order to do so,
however, we must relate our notion of quasicharacter sheaf to the sheaf side of the standard dictionary.
Choose any $\Fq$-rational point of $G$ and let $\bg$ be the geometric point on $G$ lying above $g$.
Recall that the \emph{Weil group} of $G$, which we will denote by $\W(G,\bg)$, is a subgroup of the \'etale
fundamental group defined by the following diagram:
\[
 \begin{tikzcd}
 1 \rar & \ar[equal]{d} \pi_1(\bG, \bg) \rar & \W(G,\bg) \rar \dar[hook] & \Weil{} \rar \dar[hook] & 1 \\
 1 \rar &  \pi_1(\bG, \bg) \rar & \pi_1(G,\bg) \rar & \Gal(\bFq/\Fq) \rar & 1.
 \end{tikzcd}
\]
The $\Fq$-rational point $g$ under the geometric point $\bg$ determines a splitting
$\Weil{}\to \W(G,\bg)$ of $\W(G,\bg)\to \Weil{}$.

For connected groups, the subcategories introduced in the previous two sections agree with $\QC(G)$:

\begin{proposition}\label{prop:connected}
  If $G$ is a connected commutative algebraic group over $\Fq$ then 
  the full subcategories $\QCf(G) \hookrightarrow \QCb(G) \hookrightarrow \QC(G)$ 
  appearing in Sections~\ref{sec:bounded} and \ref{sec:finite} are equivalences.
\end{proposition}

@@

  Since $G$ is connected, the geometric point $\bg$ determines
  an equivalence between the category of $\ell$-adic Weil local systems on $G$ and
  $\ell$-adic representations of $\W(G,\bg)$ \cite{deligne:80a}*{1.1.12}.
  Now let $(\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$ 
  and let $\rho : \W(G, \bg) \to \EEx$ be the character determined by $(\gqcs{L},\phi)$.
  Composing with the splitting $\Weil{} \to \W(G,\bg)$ yields an $\ell$-adic character
  $\rho_g : \Weil{} \to \EEx$, which is the same as the trace of Frobenius defined in Section~\ref{sec:Frob}:
  \[
  \rho_g(\Frob{}) =  \trFrob{\qcs{L}}(g).
  \]

  On the other hand, we have already seen that $\trFrob{\qcs{L}} : G(\Fq) \to \EEx$
  is a group homomorphism. Since $G$ is an algebraic group over $\Fq$, it is a
  variety over $\Fq$ and thus $G(\Fq)$ is finite.
  Therefore $\trFrob{\qcs{L}}(g) = \rho_g(\Frob{})$ is a root of unity
  for every $g\in G(\Fq)$.  Since $\Weil{}$ is generated by
  $\Frob{}$ and $\rho_g : \Weil{} \to \EEx$ is
  a character, it follows that the image of $\rho_g$ is a finite group.
  Thus, $\rho_g$ extends to an $\ell$-adic character of $\Gal(\bFq/\Fq)$, 
  which we will also denote $\rho_g$.

  We may now lift the $\ell$-adic character $\rho_g : \Gal(\bFq/\Fq) \to \EEx$
  to an $\ell$-adic character $\pi_1(G,\bg) \to \EEx$ using the canonical topological group homomorphism
  $\pi_1(G,\bg) \to \Gal(\bFq/\Fq)$. But $\bg$ also
  determines an equivalence between the category of $\ell$-adic
  representations of $\pi_1(G,\bg)$ and $\ell$-adic local systems on $G$. Let
  $\qcs{L}_0$ be a local system on $G$ in the isomorphism class
  determined by this $\ell$-adic character of $\pi_1(G,\bg)$.
  Then $b_G^*\qcs{L}_0 \iso \gqcs{L}$.
  Since $b_G^*$ is full and faithful
  \citelist{\cite{deligne-katz:SGA7.2}*{Expos\'e XIII} \cite{beilinson-bernstein-deligne:81a}*{Prop. 5.2.1}},
  $
  b_{G\times G}^* : \Hom(m^*\qcs{L}_0,\qcs{L}_0\boxtimes\qcs{L}_0) \to \Hom({\bar m}^*\gqcs{L},\gqcs{L}\boxtimes\gqcs{L})
  $ 
  is a bijection 
  (hom taken in the categories on constructible $\ell$-adic sheaves on 
  $G\times G$ and ${\bar G}\times {\bar G}$ respectively, 
  in which $\ell$-adic local systems sit as full subcategories). 
  Let $\mu_0 : m^*\qcs{L}_0 \to \qcs{L}_0\boxtimes\qcs{L}_0$ be the isomorphism matching 
  $\mu : {\bar m}^*\gqcs{L} \to \gqcs{L}\boxtimes\gqcs{L}$, 
  the latter appearing in the definition of $\qcs{L}$. 
  Then, as in Section~\ref{sec:bounded}, $(\qcs{L}_0,\mu_0)$ is an object in $\QCb(G)$ 
  and $\qcs{L} \ceq (\gqcs{L},\mu,\phi)$ is isomorphic to $(b_G^*\qcs{L}_0,b_{G\times G}^*\mu_0)$ in $\QC(G)$.
  Thus, the full and faithful functor $B_G^* : \QCb(G) \to \QC(G)$ from Section~\ref{sec:bounded}
  is also essentially surjective, hence an equivalence.
  %(when $G$ is a connected, commutative algebraic group over $\Fq$).

@@

  To show that $L_G(f,\psi) \iso (\qcs{L}_0,\mu_0)$ we simply apply the sheaf--function dictionary
  as follows \citelist{\cite{laumon:87a}*{1.1.3} \cite{deligne:SGA4.5}*{Sommes trig.}}.
  First, recall that $L_G(f,\psi) \ceq (L(f,\psi), \mu(f,\psi))$.
  Let $k'/k$ be any degree-$n$ extension and let $\Nm : k'\to k$ be the norm map. 
  By \cite{laumon:87a}*{1.1.3.3}, 
  \[
  \trFrob{(G_{k'}\hookrightarrow G)^*L(f,\psi)} = \trFrob{\qcs{L}} \circ \Nm;
  \]
  on the other hand, 
  \[
  \trFrob{\qcs{L}} \circ \Nm = \trFrob{(G_{k'}\hookrightarrow G)^*\qcs{L}},
  \]
   by Proposition~\ref{prop:basechange}.
  %Using notation from \cite{laumon:87a}*{1.1}, this means $t_{\bullet L(f,\psi)} = t_{\bullet \qcs{L}_0}$.
  Using \cite{laumon:87a}*{Th\'eor\`eme~1.1.2}, it follows that  $L(f,\psi) \iso \qcs{L}_0$ in $\Loc(G)$.
  It is now clear that $(L(f,\psi),\mu(f,\psi)) \iso (\qcs{L}_0,\mu_0)$ in $\QCb(G)$.
  This completes the proof that $L_G : \QCf(G) \to \QC(G)$ is essentially surjective.
  Since we have already seen (in Section~\ref{sec:finite}) that $L_G$ is full and faithful, 
  it follows that $L_G$ is an equivalence.
\end{proof}

\begin{corollary}
 If $G$ is a connected, commutative algebraic group over $\Fq$ then
 \[
  \trFrob{G} : \QCiso{G} \to G(\Fq)^*
 \]
 is an isomorphism of groups.
\end{corollary}
\begin{proof}
 It suffices to observe that the group homomorphism
 $\trFrob{G} : \QCiso{G} \to G(\Fq)^*$
 is surjective because each character $\chi \in G(\Fq)^*$ determines a finite quasicharacter
 sheaf $(\Lang,\chi^{-1})$ such that the trace of Frobenius of the quasicharacter sheaf
 $B_G^* (L_G(\Lang,\chi^{-1}))$ is $\chi$.
\end{proof}
